Testing Quasar Unification with Clumpy Wind Models

### 2015-09-20 paper.tex
did more stuff

\title
{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Unifying quasars with clumpy wind models
}

% [Matthews, J.]
% \author[Matthews, J.]{James Matthews \\
% {\sl  Supervisor: Prof. Christian Knigge} \\
% {\sl School of Physics \& Astronomy, University of Southampton,
%   Southampton, SO17 1BJ, UK}}

@@

Various unification schemes have been proposed to interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing ``average'' AGN spectra to synthetic spectra of simple biconical disc wind models, using  with our Monte Carlo radiative transfer and photoionization code {\sc Python}.
In particular, we investigate whether clumping of the outflow allows us to produce synthetic spectra in the rest-frame UV that have the characteristics of AGN, assuming total luminosities as high as 10$^{45}$ ergs s$^{-1}$. 

@@

features seen in LoBAL quasars. Despite these successes, we are unable to reproduce 

@@

% remarkably similar of line-to-continuum ratios apparent 
seen in BAL and non-BAL quasar composites. {\bf This sentence needs to be more precise, especially in light ot the statement earlier in the abstract about ``good agreement''}

@@

We briefly explore the impact of general relativistic effects
on the angular distribution of disc continuum emission, and find that 
they are not sufficient to produce a self-consistent model.
% relativistic beaming, gravitational redshift and 
% We find that these general relativistic effects are not yet sufficient to produce a
% self-consistent model. 

@@

% Introduction focussing on key points
% \begin{itemize}
% \item standard wind + BALQSO introduction
% \item focus on unification and that we are testing it
% \item some discussion of scales, referencing e.g. reverb maps, variability, microlensing, Arav 
% \item clumping background: stellar winds, clumping in AGN winds, variability
% \end{itemize}

The spectra of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is often attributed to emssion from a 
geometrically thin, optically thick accretion disc surrounding the central black hole, similar
to that described by \cite{shakurasunyaev1973}.  The lines are usually attributed to scattering and absorption of material above the disk.  {\bf Note: the first half of this paragraph and the second do not follow from one another.  The first half is about spectra the second about outflows and the connection to spectra is not made until the next paragrph.  This needs fixing.}
In addition to the inflowing accreting material, 
outflows are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
potential feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required in models of galaxy evolution \citep{springel2005}
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

Approximately $20\%$ of quasars exhibit blueshifted ($\sim 0.1c$)
broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within the ADW paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}.
ADWs offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.
Within this framework, the broad-line region (BLR) can 
correspond either to the dense wind base or dense clumps embedded
in the outflows. Indeed, \cite{elitzur2014} found that a disc-wind BLR scenario
naturally explains the emission line evolution of AGN.
A biconical wind model can also readily explain the various sub-classifications of BALQSOs: 
HiBALQSOs, which only exhibit high ionization line absorption; LoBALQSOs, which also show
absorption in lower ionization state species such as \mg\ and \al; and
FeLoBALQSOs which show further absorption in Fe~\textsc{ii} and \textsc{iii}.
In unified models, this is generally attributed to ionization stratification
of the outflow \citep[e.g.][]{elvis2000}.

@@

{\bf This needs an introductory paragraph.  Should state our earlier modeling of AGN H13 was based on a smooth outflow.  It should reference the introduction where we say there are observational indications of clumping.  Should say that the difficulty associated with clumping is two fold - computationally implies many grid cells,  many potential parameters.  Ultimately, say that for simplicity we use microclumping.  Then say this has been used elsewhere, but with more recent references than what we have.  The point is we need to motivate micro clumping before we say what it means to the code  It's all mixed up as written. Aside: at some point this paper should mention H14, and hydro models more generally}

To allow for clumping in our outflow we adopt a simple approximation
used extensively in stellar wind modelling, known as 
{\em microclumping} \citep{hamann1998}. The key assumption here is that 
typical clump sizes are much smaller than the typical photon mean free path, 
and thus the clumps are 
both geometrically and optically thin. This approach 
allows one to introduce a `filling factor', $f$, which is the 
fraction of the volume of the plasma filled by clumps. The intra-clump
medium is assumed to be a vacuum. We can then introduce the 
density enhancement, $D$, which is defined as 
\begin{equation}
D = \frac{1}{f}.
\end{equation}
We then multiply all densities in the model by $D$, and all emitting volumes
by $f$. This has the effect of enhancing all emissivities and opacities
that scale with the square of density (such as collisional excitation and recombination) 
by a factor $D$. All processes that scale linearly with density 
(such as electron scattering and bound-free opacity)
will remain unchanged for a given ionization state. 
% A factor $f$ is also
% applied to the opacities such that opacities which scale only with $\rho$ are not
% increased.

@@

Here we describe the results from our model, the parameters
of which are shown in table~1.
Parameters differing from the benchmark model of
H13 are highlighted with an asterisk.
This set of parameters
was arrived at by conducting a limited grid search over a 
5-dimensional parameter space involving the variables
$r_{min}$, $\theta_{min}$, $f$, $\alpha$ and $R_V$.
The full grid, including output spectral files and plots can be found at
\url{jhmatthews.github.io/quasar-wind-grid/}.
We evaluated these models qualitatively based on the following
criteria:
\begin{itemize}
\item Does the model maintain the right ionization state to produce strong BALs?
\item Does significant line emission emerge at low inclinations?
\item Do H recombination lines appear in the spectrum?
\item Do a certain range of angles produce LoBAL features?
\item Does the model compare favourably to quasar composite spectra?
\end{itemize}
In this section, we present one of the most promising models and discuss
the various successes and failures.
This allows us to gain insight into fundamental geometrical 
and physical constraints and assess the potential for unification.

@@

\subsubsection{The Disc SED}
\label{discsed}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/agnspec.png}
\caption
{
$F_\lambda$ for three different wavelengths
as a function of inclination from \agn\ models, 
compared to a classical AD. The \agn\ models are computed 
for Kerr and Scharzschild BHs with the same $M_{BH}$ and 
$\dot{m}$ as our model. 
}
\label{fig:f2000}
\end{figure}

When comparing BALQSO and quasar composites, it is apparent
that they possess remarkably similar line strengths and widths 
\citep[e.g.][]{weymann1991,reichard2003}.  
{\bf I think we mean EW here; otherwise I don't understand what we are talking about.  We could say ``line shapes and equivalent widths''.  I also think this requries more than one sentence of introduction.}
>>>>>>> 3c3265bb9688e4788107378165d92e06c2aa7abe
This presents a challenge to our model, as well as the geometric 
unification picture in general, because these quantities should vary with inclination angle.
Limb darkening and foreshortening causes the disc continuum emission to be 
strongly anisotropic, yet the line emission in our models is much more isotropic. 
This has the effect of enhancing the line-to-continuum ratios at high inclinations. 
To construct a scenario where emission 
line equivalent widths are comparable at all inclinations requires 
significant fine-tuning when considering a classical thin disc. {\bf Is the previous sentence a throwaway line, or has some person done this.  If so, please reference; if not please delete.}
%An alternative, simpler solution is that the emergent continuum is roughly isotropic.

Obviously,  one possible solution to this problem is hypothesize a more isotropic distribution for the emergent condition than predicted by a ``classical'' disc. 
General relativistic effects -- specifically, light bending
and relativistic beaming -- can cause 
the accretion disc SED to become more isotropic \citep[e.g.][]{zhang1997,munozdarias2013}.  Incorporating these effects into {\sc PYTHON} would be a very difficult and time-consuming challenge. Therefore, for the purposes of this discussion we have chosen to investigate whether GR effects could produce enough isotropy to address the EW problem, we have generated 
\ GR disc spectra, we use the code \agn\ \citep{hubeny2000,davishubeny2006,davis2007}. 
The output flux at three different wavelengths
as a function of inclination for an \agn\ model with the same disc and BH parameters
as our clumpy wind model is shown in figure~\ref{fig:f2000}.  The results indicate that 
the effects of GR on an AGN disc are much less extreme 
in the UV portion of the spectrum than the calculations
by \cite{zhang1997} for the X-rays in X-ray binaries.
The emission for the disks of AGN is still strongly anisotropic.
GR alone therefore cannot explain the line ratio trends in quasars by making the disc
emit more isotropically. This supports the findings of \cite{risaliti2011}, 
who find that EW distributions in quasars are consistent with anisotropic emission 
from optically thick, 
disc-like structures for {\em both} the continuum source and BLR. 
If this is the case, it has a dramatic affect on the intrinsic
BAL fraction inferred from flux-limited samples \citep{goodrich1997,krolikvoit1998}.
We will explore these ideas further in a future study.
We will explore these ideas further in a future study.  {\bf Wait!  All this says is that there has to be an alternative explanation.  It is not support for any specific explanation!. This last few sentences should be a separate paragraph.  They should start out: An alernative explanation proposed by risalite is that...}

@@

% Here are some notes on various parts of the text.
% \smallskip

% \begin{itemize}
% 	\item 
% 	\item Figure 3 should be improved to include some kind of real colormap of the model
% 	\item At the moment I haven't included another plot for the optical spectrum, or discussed
% 	it in the paper.
% 	\item Figure 5 needs to show a sample at the higher luminosity end, and include the actual relationship as the dotted line instead of a by eye version as is currently 
% 	\item
% 	\item
% 	\item
% 	\item
% 	\item
% 	\item
% 	\item
% \end{itemize}

>>>>

\title
{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Testing Quasar Unification with Clumpy Wind Models
%Unifying quasars with clumpy wind models
}

@@

Various unification schemes have been proposed to interpret the complex phenomenology 
of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric 
picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing ``average'' AGN spectra to 
synthetic spectra of simple biconical disc wind models, using our state-of-the-art 
Monte Carlo radiative transfer and photoionization code.
In particular, we investigate whether clumping of the outflow allows us to 
produce synthetic spectra in the rest-frame UV that have the characteristics of AGN, 
assuming total luminosities as high as 10$^{45}$ ergs s$^{-1}$. 

@@

features seen in LoBAL quasars. 
Despite these successes, we are unable to reproduce 

@@

seen in BAL and non-BAL quasar composites. 
{\bf This sentence needs to be more precise, especially in 
light of the statement earlier in the abstract about ``good agreement''}

@@

@@

The spectra of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is often attributed to emssion from a 
geometrically thin, optically thick accretion disc surrounding the central black hole, similar
to that described by \cite{shakurasunyaev1973}.  The lines are usually attributed to scattering and absorption of material above the disk.  
{\bf Note: the first half of this paragraph and the second do not follow from one another.  
The first half is about spectra the second about outflows and the connection to spectra is not made until the next paragrph.  
This needs fixing. JM: Knox, I like this structure...I start with inflow, then move to outflow.}
In addition to the inflowing accreting material, 
outflows are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
potential feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required in models of galaxy evolution \citep{springel2005}
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

Approximately $20\%$ of quasars exhibit blueshifted ($\sim 0.1c$)
broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within the ADW paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}, although this geometry is contradicted
by more recent studies of radio spectral indices (REFS DIPOMPEO; see section ?).
ADWs offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.
Within this framework, the broad-line region (BLR) can 
correspond either to the dense wind base or dense clumps embedded
in the outflows. Indeed, \cite{elitzur2014} show that a disc-wind BLR scenario
naturally explains the emission line evolution of AGN.
A biconical wind model can also readily explain the various sub-classifications of BALQSOs: 
HiBALQSOs, which only exhibit high ionization line absorption; LoBALQSOs, which also show
absorption in lower ionization state species such as \mg\ and \al; and
FeLoBALQSOs which show further absorption in Fe~\textsc{ii} and \textsc{iii}.
In unified geometric models, this is generally attributed to ionization stratification
of the outflow \citep[e.g.][]{elvis2000}.

@@

% {\bf This needs an introductory paragraph.  Should state our earlier modeling of AGN H13 was based on a smooth outflow.  It should reference the introduction where we say there are observational indications of clumping.  Should say that the difficulty associated with clumping is two fold - computationally implies many grid cells,  many potential parameters.  Ultimately, say that for simplicity we use microclumping.  Then say this has been used elsewhere, but with more recent references than what we have.  The point is we need to motivate micro clumping before we say what it means to the code  It's all mixed up as written. Aside: at some point this paper should mention H14, and hydro models more generally}

To allow for clumping in our outflow we adopt a simple approximation
used extensively in stellar wind modelling, known as 
{\em microclumping} \citep{hamann1998}(MORE REFS). The key assumption here is that 
typical clump sizes are much smaller than the typical photon mean free path, 
and thus the clumps are 
both geometrically and optically thin. This approach 
allows one to introduce a `filling factor', $f$, which is the 
fraction of the volume of the plasma filled by clumps. The intra-clump
medium is assumed to be a vacuum. We can then introduce the 
density enhancement, $D$, which is defined as 
\begin{equation}
D = \frac{1}{f}.
\end{equation}
We then multiply all densities in the model by $D$, and all emitting volumes
by $f$. This has the effect of enhancing all emissivities and opacities
that scale with the square of density (such as collisional excitation and recombination) 
by a factor $D$. All processes that scale linearly with density 
(such as electron scattering and bound-free opacity)
will remain unchanged for a given ionization state. 
% A factor $f$ is also
% applied to the opacities such that opacities which scale only with $\rho$ are not
% increased.

@@

Here we describe the results from our model, the parameters
of which are shown in table~1.
Parameters differing from the benchmark model of
H13 are highlighted with an asterisk.
This set of parameters
was arrived at by conducting a limited grid search over a 
5-dimensional parameter space involving the variables
$r_{min}$, $\theta_{min}$, $f$, $\alpha$ and $R_V$.
We evaluated these models based on the following
criteria:
\begin{itemize}
\item Does the model maintain the right ionization state to produce strong BALs ($BI > ??$)?
\item Does significant line emission emerge at low inclinations, with $EW\sim50\AA$ in \civ?
\item Do H recombination lines appear in the spectrum, $EW\sim50\AA$ in \la?
\item Do a certain range of angles produce LoBAL features?
\item Does the model compare favourably to quasar composite spectra?
\end{itemize}
In this section, we present one of the most promising models and discuss
the various successes and failures.
This allows us to gain insight into fundamental geometrical 
and physical constraints and assess the potential for unification. 
The full grid, including output synthetic spectra and plots can be found at
\url{jhmatthews.github.io/quasar-wind-grid/}.

@@

\subsubsection{The Angular Distribution of Line And Continuum Emission}
\label{discsed}

@@

When comparing BALQSO and quasar composites, it is apparent
that they possess remarkably similar line strengths and widths 
\citep[e.g.][]{weymann1991,reichard2003}.  
{\bf I think we mean EW here; otherwise I don't understand what we are talking about.  We could say ``line shapes and equivalent widths''.  I also think this requries more than one sentence of introduction.}
This presents a challenge to our model, as well as the geometric 
unification picture in general, because these quantities should vary with inclination angle.
Limb darkening and foreshortening causes the disc continuum emission to be 
strongly anisotropic, yet the line emission in our models is much more isotropic. 
This has the effect of enhancing the line-to-continuum ratios at high inclinations. 
To construct a scenario where emission 
line equivalent widths are comparable at all inclinations requires 
significant fine-tuning when considering a classical thin disc. {\bf Is the previous sentence a throwaway line, or has some person done this.  If so, please reference; if not please delete.}
%An alternative, simpler solution is that the emergent continuum is roughly isotropic.

Obviously,  one possible solution to this problem is hypothesize a more isotropic distribution for the emergent condition than predicted by a ``classical'' disc. 
General relativistic effects -- specifically, light bending
and relativistic beaming -- can cause 
the accretion disc SED to become more isotropic \citep[e.g.][]{zhang1997,munozdarias2013}.  Incorporating these effects into {\sc PYTHON} would be a very difficult and time-consuming challenge. Therefore, for the purposes of this discussion we have chosen to investigate whether GR effects could produce enough isotropy to address the EW problem, we have generated 
\ GR disc spectra, we use the code \agn\ \citep{hubeny2000,davishubeny2006,davis2007}. 
The output flux at three different wavelengths
as a function of inclination for an \agn\ model with the same disc and BH parameters
as our clumpy wind model is shown in figure~\ref{fig:f2000}.  The results indicate that 
the effects of GR on an AGN disc are much less extreme 
in the UV portion of the spectrum than the calculations
by \cite{zhang1997} for the X-rays in X-ray binaries.
The emission for the disks of AGN is still strongly anisotropic.
GR alone therefore cannot explain the line ratio trends in quasars by making the disc
emit more isotropically. This supports the findings of \cite{risaliti2011}, 
who find that EW distributions in quasars are consistent with anisotropic emission 
from optically thick, 
disc-like structures for {\em both} the continuum source and BLR. 
If this is the case, it has a dramatic affect on the intrinsic
BAL fraction inferred from flux-limited samples \citep{goodrich1997,krolikvoit1998}.
We will explore these ideas further in a future study.  
{\bf Wait!  All this says is that there has to be an alternative explanation.
It is not support for any specific explanation!. This last few sentences should
be a separate paragraph.  They should start out: An alernative explanation proposed 
by risalite is that...}

@@
